---
title: "Independent Study to consolidate this week"
subtitle: "Two-way ANOVA"
toc: true
toc-location: right
format:
  html:
    code-fold: true
    code-summary: "Answer - don't look until you have tried!"
---

# Set up

If you have just opened RStudio you will want to load the
**`tidyverse`** and **`emmeans`** packages.

```{r}
#| code-fold: false
library(tidyverse)
library(emmeans)
```

# Exercises

1.  💻 Researchers suspect there may be regional differences between two
    species of butterfly, *F.concocti* and *F.flappa*. They catch 8
    randomly chosen individuals of each species in two regions (North
    and South) and measure their wing lengths in mm. The data are in
    [butterf.txt](data-raw/butterf.txt). What do you conclude about the
    size of these species in the North and South?

```{r}
#| output: false

butter  <-  read_table("data-raw/butterf.txt")
str(butter)

```

```{r}
#| output: false

# create a rough plot of the data  
ggplot(data = butter,
       aes(x = region, y = winglen, fill = spp)) +
  geom_violin()
# There seems to be little difference between the two species in the
# south but F.concocti
# is larger than F.flappa in the North.
```

```{r}
#| output: false

# create a summary of the data
butter_summary <- butter |> 
  group_by(region, spp) |> 
  summarise(mean = mean(winglen),
            median = median(winglen),
            sd = sd(winglen),
            n = length(winglen),
            se = sd/sqrt(n))
```

```{r}
#| output: false

# The data seem to be continuous so it is likely that a parametric
# test will be fine
# we will check the other assumptions after we have run the lm

# build the statistical model
mod <- lm(data = butter, winglen ~ spp * region)


# examine it
summary(mod)


# the line starting (Intercept) is  β0  
# the line starting sppF.flappa is β1
# the line starting regionsouth is β2
# the line starting sppF.flappa:regionsouth is β3

# F.concocti-north mean is β0 i.e., 32.275
# F.Flappa-north mean is β0 + β1 i.e., 32.275 - 7.875 = 24.4
# F.concocti-south is β0 + β2  i.e., 32.275 - 8.638 = 23.637
# F.Flappa--south is β0 + β1 + β2 + β3 
#           i.e., 32.275 - 7.875 - 8.638 + 7.713 = 23.475

# The model of spp and region overall explains a significant 
# amount (49%) of the variation in wing lengths
# (F = 8.1; d.f. = 3, 28; p = 0.0002). To see which of the three 
# effects are significant we can use the `anova()` function on our
# model.
```

```{r}
#| output: false
anova(mod)

# There was a significant effect of species (F = 8.1; 
# d.f. = 1, 28; p = 0.008) and region (F = 11.5; 
# d.f. = 1, 28; p = 0.002) on wing length and these effects interact 
# (F = 7.5; d.f. = 1, 28; p = 0.01)

# It is important to consider how the interaction effect is
# influencing our interpretation of the main effects.
# Judging by the means and the rough plot, the significant
# effect of both species and region is due to the fact that
# F.concocti is larger than F.flappa in the North.
# In other words, we only have a significant effect of species
# in the North. 


# the emmeans package will let us see exactly what comparisons
# are significant
```

```{r}
#| output: false
# post-hoc comparisons
emmeans(mod, ~ spp * region) |> pairs()

# As we suspected all the difference are between  F.concocti north
# and the other three groups. 
 # F.concocti north - F.flappa north    p = 0.0026
 # F.concocti north - F.concocti south  p = 0.0010
 # F.concocti north - F.flappa south    p = 0.0008

# There was a signifcant interaction between the effects of species 
# and region (F = 7.5; d.f. = 1, 28; p = 0.01) with Northern
# F.concocti being significantly larger than F.concocti in the 
# south (Tukey HSD: p = 0.0010) and F.flappa in the 
# north (p = 0.0026) and south (p = 0.0008). This creates a 
# significant difference between species averaged over region 
# (F = 8.1; d.f. = 1, 28; p = 0.008) and region averaged over species
```

```{r}
#| output: false
# extractpvalues
ph_tidy <- emmeans(mod, ~ spp * region) |>
  pairs() |>
  broom::tidy()

p_concn_v_flapn <- ph_tidy |> 
  filter(contrast == "F.concocti north - F.flappa north") |> 
  pull(adj.p.value) |> round(4)
p_concn_v_concs <- ph_tidy |> 
  filter(contrast == "F.concocti north - F.concocti south") |> 
  pull(adj.p.value) |> round(4)
p_concn_v_flaps <- ph_tidy |> 
  filter(contrast == "F.concocti north - F.flappa south") |> 
  pull(adj.p.value) |> round(4)

```

```{r}
#| output: false

# let's check the assumptions
plot(mod, which = 1) 
# we're looking for the variance in the residuals to be the same in both groups.
# This looks OK. Maybe a bit higher in the wild plants (with the higher mean)
 
hist(mod$residuals)
shapiro.test(mod$residuals)
# On balance the use of lm() is probably justifiable  The variance isn't quite equal 
# and the histogram looks a bit off normal but the normality test is NS and the 
# effect (in the figure) is clear.
```

```{r}
#| output: false

# A figure 
butter_fig <- ggplot() +
  geom_point(data = butter, 
             aes(x = region,
                 y = winglen,
                 shape = spp),
             position = position_jitterdodge(dodge.width = 1,
                                             jitter.width = 0.2,
                                             jitter.height = 0),
             size = 2) +
  geom_errorbar(data = butter_summary, 
                aes(x = region, 
                    ymin = mean - se, 
                    ymax = mean + se, 
                    group = spp),
                width = 0.4, 
                position = position_dodge(width = 1)) +
  geom_errorbar(data = butter_summary, 
                aes(x = region, 
                    ymin = mean, 
                    ymax = mean, 
                    group = spp),
                width = 0.3, 
                position = position_dodge(width = 1) ) +
  scale_x_discrete(name = "region") +
  scale_y_continuous(name = "Wing length (mm)",
                     expand = c(0, 0),
                     limits = c(0, 55)) +
  scale_shape_manual(values = c(19, 1),
                     name = NULL,
                     labels = c(bquote(italic("F.concocti")),
                                bquote(italic("F.flappa")))) +
  # F.concocti north - F.flappa north    p = 0.0026
  annotate("segment",
           x = 0.75, xend = 1.25,
           y = 40, yend = 40,
           colour = "black") +
  annotate("text",
           x = 1,  y = 42,
           label = glue::glue("italic(p) == { p_concn_v_flapn }"),
           parse = TRUE) +
  # F.concocti north - F.concocti south  p = 0.0010
  annotate("segment",
           x = 0.75, xend = 1.75,
           y = 44, yend = 44,
           colour = "black") +
  annotate("text", x = 1.25,  y = 46,
           label = glue::glue("italic(p) == { p_concn_v_concs }"),
           parse = TRUE) +
  # F.concocti north - F.flappa south    p = 0.0008
  annotate("segment",
           x = 0.75, xend = 2.25,
           y = 48, yend = 48,
           colour = "black") +
  annotate("text", x = 1.5,  y = 50,
           label = glue::glue("italic(p) == { p_concn_v_flaps }"),
           parse = TRUE) +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "inside",
        legend.position.inside = c(0.2, 0.12),
        legend.background = element_rect(colour = "black"))


# save figure to figures/butter.png
ggsave("figures/butter.png",
       plot = butter_fig,
       width = 3.5,
       height = 3.5,
       units = "in",
       dpi = 300)

```
