---
title: "Independent Study to consolidate this week"
subtitle: "Summarising data in with several variables and the role of variables in analysis"
toc: true
toc-location: right
format:
  html:
    code-fold: true
    code-summary: "Answer - don't look until you have tried!"
---

# Set up

If you have just opened RStudio you will want to load the packages and import the data.

```{r}
#| code-fold: false
library(tidyverse)
library(readxl)
```


1.  💻 Summarise and plot the `pigeons` dataframe appropriately.

```{r}
#| fig-keep: none

# import
pigeons <- read_table("data-raw/pigeon.txt")

# reformat to tidy
pigeons <- pivot_longer(data = pigeons, 
                        cols = everything(), 
                        names_to = "population", 
                        values_to = "distance")

# sumnmarise
pigeons_summary <- pigeons %>%
  group_by(population) %>%
  summarise(mean = mean(distance),
            std = sd(distance),
            n = length(distance),
            se = std/sqrt(n))
# plot
ggplot() +
  geom_point(data = pigeons, aes(x = population, y = distance),
             position = position_jitter(width = 0.1, height = 0),
             colour = "gray50") +
  geom_errorbar(data = pigeons_summary, 
                aes(x = population, ymin = mean - se, ymax = mean + se),
                width = 0.3) +
  geom_errorbar(data = pigeons_summary, 
                aes(x = population, ymin = mean, ymax = mean),
                width = 0.2) +
  scale_y_continuous(name = "Interorbital distance (mm)", 
                     limits = c(0, 14), 
                     expand = c(0, 0)) +
  scale_x_discrete(name = "Population") +
  theme_classic()

```


2. 💻 The data in [blood.csv](data-raw/blood.csv) are measurements of several blood parameters from fifty people with [Crohn's disease](https://en.wikipedia.org/wiki/Crohn%27s_disease), a lifelong condition where parts of the digestive system become inflamed. Twenty-five of people are in the early stages of diagnosis and 25 have started treatment. The variables in the dataset are:

    -   sodium  - Sodium concentration in umol/L, the average of 5 technical replicates
    -   potassium - Potassium concentration in umol/L, the average of 5 technical replicates
    -   B12 Vitamin - B12 in pmol/L, the average of 5 technical replicates
    -   wbc - White blood cell count in 10^9 /L, the average of 5 technical replicates
    -   rbc count - Red blood cell count in	10^12 /L, the average of 5 technical replicates
    -   platlet count - platlet count in	10^9 /L, the average of 5 technical replicates
    -   inflammation marker - the presence or absence of a marker of inflammation, either 0 or 1 
    -   status - whether the individual is before or after treatment.



    Your task is to summarise and plot these data in any suitable way. Create a complete RStudio Project for an analysis of these data. You will need to:

    -   Make a new project 
    -   Make folders for data and for figures
    -   Import the data
    -   Summarise and plot variables of your choice. It doesn't matter what you chose - the goal is the practice the project workflow and selecting appropriate plotting and summarising methods for particular data sets.