---
title: "Independent Study - after the workshop"
execute: 
  echo: false
  include: false
---

### Blood data set

```{r}
# set up
# pkgs
library(tidyverse)
library(readxl)

```

1.  💻 Summarise the `fly_bristles_means` dataframe by calculating the mean, median, sample size, standard deviation and standard error of the `mean_count` variable.

```{r}
#| code-fold: true
fly_bristles_means |> 
  summarise(mean = mean(mean_count),
            median = median(mean_count),
            n = length(mean_count),
            standard_dev = sd(mean_count),
            standard_error = standard_dev / sqrt(n))
```

2.  💻 Create an appropriate plot to show the distribution of `mean_count` in `fly_bristles_means`

```{r}
#| code-fold: true

ggplot(fly_bristles_means, aes(x = mean_count)) +
  geom_histogram(bins = 10)
```

3.  💻 Can you format the plot 2. by removing the grey background, giving the bars a black outline and the fill colour of your choice and improving the axis format and labelling? You may want to refer to [last week's workshop](r4babs1/week-7/workshop.html).

```{r}
#| code-fold: true

ggplot(fly_bristles_means, aes(x = mean_count)) +
  geom_histogram(bins = 10, 
                 colour = "black",
                 fill = "skyblue") +
  scale_x_continuous(name = "Number of bristles",
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Frequency",
                     expand = c(0, 0),
                     limits = c(0, 35)) +
  theme_classic()
```

