---
title: "Workshop"
subtitle: "Data Analysis 2: Biomedical sciences - Sample data analysis"
toc: true
toc-location: right
---

```{r}
#| include: false
library(kableExtra)
```

# Introduction

## Session overview

In this workshop you will learn how to analyse some sample flow
cytometry data which is formatted in the same way as the data you will
produce.

# Exercises

## Set up

### Make a RStudio Project

ðŸŽ¬ Start RStudio from the Start menu.

ðŸŽ¬ Make an RStudio project. Be deliberate about where you create it so
that it is a good place for you.

ðŸŽ¬ Use the Files pane to make a folders for the data. I suggest
`data-raw/`, `data-processed/` and `data-meta/`

ðŸŽ¬ Make a new script called `sample-data-analysis.R` to carry out the
rest of the work.

ðŸŽ¬ Record what you do and what you find out. All of it! Make notes on
the inputs and outputs of each command. This will help you work out
problems analysing your own data.

### Load packages

We need three packages for this workshop:

-   **`tidyverse`** [@tidyverse]: importing the meta data which is in a
    text file, working with the data once it is in a dataframe to
    filter, summarise and plot.

-   **`flowCore`** [@flowCore]: to import data from flow cytometry files
    (`.fcs`), apply the logicle transformation, and use functions for
    using a "flowSet" data structure.

-   **`flowAI`** [@flowAI]: which performs automated quality control by
    checking for changes in instrument speed and signal intensity that
    indicate a problem for the reading for that cells

::: callout-tip
## Working on University computers or using the Virtual Desktop Service

These packages are installed already. You can go straight to loading
them.
:::

::: callout-warning
## Working on your own machine

I recommend working on the University computers for this work. You can
still work from home by using the [Virtual Desktop
Service](https://www.york.ac.uk/it-services/tools/vds/). The VDS allows
you to log on to a university computer from your own computer. It means
you can access all software and filestores. When using the VDS for R a
nd RStudio, it usually makes sense to use other software - such as a
browser or file explorer - also through the VDS. Go to the [Virtual
Desktop Service](https://www.york.ac.uk/it-services/tools/vds/) for set
up instructions.

However, If you are confident in your ability to set up your own
machine, you will need to install these packages.

You can install **`tidyverse`** from CRAN in the normal way.

**`flowCore`** and **`flowAI`** are not on CRAN but come from
[Bioconductor](https://www.bioconductor.org/).

-   first install **`BiocManager`** from CRAN in the normal way.

-   then install from Bioconductor using
    `BiocManager::install("flowCore")` and
    `BiocManager::install("flowAI")`

If you have difficulty installing these packages, use the [Virtual
Desktop Service](https://www.york.ac.uk/it-services/services/vds/) which
allows you to log into a university machine from your laptop.
:::

ðŸŽ¬ Load **`tidyverse`**,**`flowCore`** and **`flowAI`**

```{r}
library(flowCore)
library(flowAI)
library(tidyverse)
```

::: callout-important
## Order of package loading

Load tidyverse last because **`flowCore`** also has some functions with
the same names as functions in **`tidyverse`** and you will want the
tidyverse ones.
:::

## Import the data

There are six data files, one for each treatment and antibody
combination:

-   [ECOLIGreen_ISOTYPE.fcs](data-raw/ECOLIGreen_ISOTYPE.fcs)
-   [ECOLIGreen_TNFAPC.fcs](data-raw/ECOLIGreen_TNFAPC.fcs)
-   [LPS_ISOTYPE.fcs](data-raw/LPS_ISOTYPE.fcs)
-   [LPS_TNFAPC.fcs](data-raw/LPS_TNFAPC.fcs)
-   [MEDIA_ISOTYPE.fcs](data-raw/MEDIA_ISOTYPE.fcs)
-   [MEDIA_TNFAPC.fcs](data-raw/MEDIA_TNFAPC.fcs)

ðŸŽ¬ Save copies of each file (right click and save as) to your
`data-raw/` folder.

These are six files associated with one experiment.

We are going to use the command `read.flowSet()` from the **`flowCore`**
package to import the data. In R, the data from one file is stored in a
"flowFrame" object. A flowFrame is a data structure in the same way that
a dataframe or a vector is a data structure. A flowFrame has "slots".

-   The first slot is called `exprs` and contains the data itself in
    matrix which looks just like a dataframe.
-   The second is called `parameters` and contains the names of the
    columns in `exprs`.
-   The third is called `description` and contains information about the
    `.fcs` file.

A "flowSet" is a list of related flowFrames. They are related in that
they have the same column names, came from the same machine, and are
part of the same experiment. Such data needs to be treated in the same
way which is why it is useful to have a structure like a flowSet.

ðŸŽ¬ Make a variable to hold the file names:

```{r}
myfiles <- list.files("data-raw", pattern = ".fcs$")
```

`list.files()` will return a character vector of the names of the files
`data-raw/`. The `pattern` argument is a "regular expression" which is
used to filter the names of the files. Here we are asking for the names
of the files which end in `.fcs`. This would be very useful if you had
other file types in that folder. In our case, there are only `.fcs`
files in that folder.

ðŸŽ¬ Type `myfiles` to check it contains what you expect.

ðŸŽ¬ Use `read.flowSet()` to import the files from the folder `data-raw/`
that are listed in the variable `myfiles` and name the resulting flowSet
`fs`:

```{r}
fs <- read.flowSet(myfiles, 
                   path = "data-raw")
```

You now have a flowSet which is a list of 6 flowFrames, one for each
`.fcs` file.

## Explore the data structure

ðŸŽ¬ Type `fs` to see what it contains:

```{r}
fs 
```

We can index the list to access the first flowFrame in the flowSet.

ðŸŽ¬ Type `fs[[1]]` to see the information about the first flowFrame in
the flowSet:

```{r}
fs[[1]] 
```

This tells you there are 50000 cells (rows) and 16 columns in this
dataset. The first two columns `TIME` and `Time MSW` give the time in
binary format since the beginning of the data acquisition.

ðŸŽ¬ You can use the `exprs()` function to access the actual data and pipe
it to `View()` to show it in a window:

```{r}
#| eval: false
exprs(fs[[1]]) |> View()
```

ðŸŽ¬ List the column names in each of the flowFrames:

```{r}
colnames(fs)
```

The side scatter (`SS`) and forward scatter (`FS`) columns are used to
select the live cells. The fluorescence measurements are in the FL
columns and it is the linear values, `FL 1 Lin` and `FL 8 Lin` that have
our FITC and APC signals respectively.

::: callout-tip
These exploratory steps are to help you understand the R objects you
have. They are not essential parts of the analysis.
:::

## Improve the column names

Names like `FL 1 Lin` meaning "fluorescence channel 1 linear" are not 
very informative. A list of more useful column names are given in
[meta.csv](data-meta/meta.csv). We can import this file and use the
resulting dataframe to rename columns in our flowFrames. This will make
it easier to understand our data and results.

ðŸŽ¬ Save a copy of the file (right click and save as) to your
`data-meta/` folder

ðŸŽ¬ Import the meta data and examine it:

```{r}
meta <- read_csv("data-meta/meta.csv")

```

```{r}
#| echo: false
meta |> 
  knitr::kable() |> 
  kable_styling() |> 
  scroll_box(height = "200px")

```

ðŸŽ¬ Assign the names in `meta$name` to the columns in the flowFrames:

```{r}
colnames(fs) <- meta$name
# view the effect
colnames(fs) 
```

::: callout-tip
You don't have to use the names I have used for your own analysis if
you prefer different names. Simply edit the `meta.csv` file. However,
you do need to keep the Time column named that way for the next step
to proceed.
:::


## Quality control 1: Automated instrument issues

**`flowAI`** is a package that provides a set of tools for automated QC
of flow cytometry data. It checks for:

-   changes in instrument speed using the Time column,
-   changes signal intensity
-   cells that are outside the dynamic range

It generates a new flowSet with the problematic events removed along
with a folder containing filtered `.fcs` files and a quality control
report in html format. It will take a few minutes to do all files.

ðŸŽ¬ Run the automated quality control:

```{r}
#| echo: false
fs_clean <- flow_auto_qc(fs, 
                         html_report = FALSE,
                         folder_results = "sample-QC")
```

```{r}
#| eval: false
fs_clean <- flow_auto_qc(fs, 
                         folder_results = "sample-QC")
```

â“ What has been removed through the QC process?

â“ What is in the folder "sample-QC"?

### Get a feel of your data: Plot distributions

Plotting the distributions of our variables is very useful for getting
an overview of our data. We'll just look at the first flowFrame to get
over the principles. However, you may wish to examine all of the flow
frames of your own data. 

ðŸŽ¬ Plot the distribution of the linear TNFa_APC signal in the first 
flowFrame in the flowSet:

```{r}
exprs(fs_clean[[1]]) |> 
  data.frame() |> 
  ggplot(aes(x = TNFa_APC_Lin)) +
  geom_histogram(bins = 100)

```

This is a very skewed distribution which makes visualisation hard. It is
common to log skewed distributions to improve visualisation. 

ðŸŽ¬ Plot the distribution of the logged `TNFa_APC_Lin` column:

```{r}
exprs(fs_clean[[1]]) |> 
  data.frame() |> 
  ggplot(aes(x = log(TNFa_APC_Lin))) +
  geom_histogram(bins = 100)

```

Now our data are easier to see.

## Transform the data

In fact, we will apply a different transformation called "logicle"
[@parks2006]. This transformation is has a similar effect as logging but
avoids some problems that can occur with flow cytometry data and
especially flow cytometry data that have been "compensated" .
Compensation is routinely applied to flow cytometry data to correct for
fluorophores being detected in in multiple channels. This was not a
problem in our experiment.

The transformation is applied in two steps: the transformation needed 
is estimated from the data and then that transformation is applied.

We need to apply the transformation only to the `TNFa_APC_Lin` and
`E_coli_FITC_Lin` columns. The `estimateLogicle()` function refers to 
columns as channels.

ðŸŽ¬ Estimate the transformation from the data:

```{r}
trans <- estimateLogicle(fs_clean[[1]],
                         channels = c("E_coli_FITC_Lin", "TNFa_APC_Lin"))
```

ðŸŽ¬ Apply the transformation to all the`TNFa_APC_Lin` and
`E_coli_FITC_Lin` columns the flowSet:

```{r}
# apply the transformation
fs_clean_trans <- transform(fs_clean, trans)
```

ðŸŽ¬ Examine the effect on the first flow frame:

```{r}
exprs(fs_clean_trans[[1]]) |> 
  data.frame() |> 
  ggplot(aes(x = TNFa_APC_Lin)) +
  geom_histogram(bins = 100)

```



That looks better.

## Make data easier to work with

We are going to get the data from the flowSet and put in a dataframe.
FlowCore contains many functions for gating (filtering), plotting and
summarising flowSets but putting the data into a dataframe will make it
easier for you to use tools you already know and more easily follow
the priciples of what you are doing.

We will be using familiar tools like `group_by()` and `summarise()`,
`filter()` and `ggplot()`

ðŸŽ¬ Put the transformed data into a data frame:

```{r}
# Put into a data frame for ease of use
clean_trans <- fsApply(fs_clean_trans, exprs) |> data.frame() 
```

`fsApply(flowSet, function)` applies a function to every flowFrame in a
flowSet. Here we have accessed the data matirx in each with `exprs()`.

The are `r nrow(clean_trans)` rows (cells) in the dataset. You can view 
it by clicking on the dataframe in the Environment. At the moment, 
we cannot tell which row (cell) is from which sample but we can add the 
sample names as a column in that dataframe.

-   Each sample name needs to appear as many times as there are cells
    (events) in the corresponding flowFrame. For example,
    `dim(fs_clean_trans[[1]])["events"]` is 
    `r dim(fs_clean_trans[[1]])["events"]` cells.

-   `sampleNames(fs_clean_trans)` gives the names of the samples.

-   We can use the `rep()` function to repeat the sample names the
    correct number of times.

ðŸŽ¬ Add the sample name to each row:

```{r}
clean_trans <- clean_trans |> 
  dplyr::mutate(sample = rep(sampleNames(fs_clean_trans),
                             times = c(dim(fs_clean_trans[[1]])["events"],
                                       dim(fs_clean_trans[[2]])["events"],
                                       dim(fs_clean_trans[[3]])["events"],
                                       dim(fs_clean_trans[[4]])["events"],
                                       dim(fs_clean_trans[[5]])["events"],
                                       dim(fs_clean_trans[[6]])["events"])))
```

The sample name contains the information about which antibody and
treatment the sample was. Our analysis will be a little easier if we add
this information to separate columns. We can extract the antibody and
treatment from the sample name using the `extract()` function and two
"regular expressions".

ðŸŽ¬ Add columns for treatment and antibody by extracting that information
from the sample name:

```{r}
clean_trans <- clean_trans |> 
  extract(sample, 
          remove = FALSE,
          c("treatment", "antibody"),
          "([a-zA-Z]+)_([a-zA-Z]+).fcs")
```

-   the sample name is treatment_antibody
-   each pattern matching the treatment and antibody is enclosed in
    parentheses
-   we want to keep those patterns to go in the new columns
-   the underscore is matched but not enclosed in parentheses so it is
    not kept
-   `.fcs` is matched but not enclosed in parentheses so it is not kept
-   the `remove = FALSE` argument means that the original column is kept
-   `[]` enclose a set of characters
-   `a-z` means any lower case letter, `A-Z` means any upper case letter
    so `[a-zA-Z]` means any letter
-   `+` means one or more of the preceding character set
-   so`[a-zA-Z]+` means one or more letters

::: callout-important
## Adding the treatment and antibody information

We have added the treatment and antibody information to each row by
processing the names of the `fcs` files. This works because the file 
names have a specific format which we have matched with the regex.
When you work with your own data, the easist thing to do is ensure your
files have the same names as the same data.
:::


ðŸŽ¬ View the dataframe to see the new columns.

Our treatments have an order. Media is the control, LPS should be next
and ECOLIGreen should be last. We can use the `fct_relevel()` function
to put groups in order so that our graphs are better to interpret.

```{r}
clean_trans <- clean_trans |> 
  mutate(treatment = fct_relevel(treatment, c("MEDIA",
                                              "LPS",
                                              "ECOLIGreen")))
```




## Save the data

We have cleaned and transform our data. It is a good idea to save it at
this point so that we can start from here if we need to.

ðŸŽ¬ Save the data:

```{r}
write_csv(clean_trans, "data-processed/ai_clean_logicle_trans.csv")

```

## Quality control 2: 'Gating' to Removing debris

Some of the cells will have died and broken during the experiment. We
need to perform quality control on our data to remove observations that
have that have the size and granularity that is not typical of a live
cell. This is done by filtering on the forward scatter (size) and side
scatter (granularity) channels.

In flow cytometry, it is common to describe filtering as "gating" and
the observations (cells) being kept as "in the gate".

We create a scatter plot of the forward scatter (size) and side scatter
(granularity) for each sample to see where we should put the gate.

We could use `geom_point()` but many points will overlap so it is
difficult to see the density of points. `geom_hex()` puts the points in
bins and colours the bin with the number of points.

ðŸŽ¬ Plot the forward scatter and side scatter for each sample

```{r}
ggplot(clean_trans, aes(x = FS_Lin, y = SS_Lin)) +
  geom_hex(bins = 128) +
  scale_fill_viridis_c() +
  facet_grid(antibody ~ treatment) +
  theme_bw()
```

We can see a cloud of points in the middle of the plot. This is where
the live cells are. Points towards the corners, especially in the bottom
left corner are likely to be cell debris. We can draw a gate around the 
cloud of points to remove the debris. We should use the same filter/gate
on all six data sets for consistency. The gates can be a rectangle, 
polygon, or ellipse. We will use a rectangle.

### Manual rectangular gate

Use the zoom on the plot window to better see the cloud of points and
the axis values. You need to use minimum and maximum $x$ and $y$ values
to define the rectangle. 

ðŸŽ¬ Define the minimum and maximum $x$ and $y$ values for the gate:

```{r}
xmin <- 15000
xmax <- 35000
ymin <- 7500
ymax <- 28000
```


::: callout-important
## Choosing values for your own data

I chose theses values but you might judge the gate differently.
You will certainly need to use different values for your own data.
The next plot will help you refine them.
:::


ðŸŽ¬ Put those values in a dataframe so we can plot the gate on the hexbin
plot:

```{r}
box <- data.frame(x = c(xmin, xmin, xmax, xmax),
                  y = c(ymin, ymax, ymax, ymin))
```

ðŸŽ¬ Plot the forward scatter and side scatter for each sample with the
gate:

```{r}
ggplot(clean_trans, aes(x = FS_Lin, y = SS_Lin)) +
  geom_hex(bins = 128) +
  scale_fill_viridis_c() +
  geom_polygon(data = box, aes(x = x, y = y), 
               fill = NA, 
               color = "red",
               linewidth = 1) +
  facet_grid(antibody ~ treatment) +
  theme_bw()
```

ðŸŽ¬ Adjust the values of `xmin`, `xmax`, `ymin`, and `ymax` to get a gate
that you are happy with.

We have drawn the gate on the plot but we need to use it to filter out
the debris (rows) from our data. We want only the cells with `FS_Lin`
values that are between the `xmin` and `xmax` and `SS_Lin` that are
between `ymin` and `ymax` values we chose.

ðŸŽ¬ Filter the data to remove the debris:

```{r}
# filter out the debris
clean_trans_live <- clean_trans |> 
  filter(between(FS_Lin, xmin, xmax),
         between(SS_Lin, ymin, ymax)) 

```

`clean_trans_live` now has `r nrow(clean_trans_live)` cells. 
We should report how many cells were *in the gate* for each sample.

ðŸŽ¬ Find the number of cells in each sample after flowAI cleaning:

```{r}
# 
clean_trans_n <-  clean_trans |> 
  group_by(antibody, treatment) |> 
  summarise(n = n()) 
```

Note than we use the clean and transformed data for summarising the
number in each set after cleaning.

```{r}
#| echo: false
knitr::kable(clean_trans_n)
```

ðŸŽ¬ Find the number of cells in each sample after flowAI cleaning *and*
removing debris

```{r}
# number of cells in each sample
clean_trans_live_n <-  clean_trans_live |> 
  group_by(antibody, treatment) |> 
  summarise(n_live = n()) 
```

Now we use the clean, transformed and gated data.


```{r}
#| echo: false
knitr::kable(clean_trans_live_n)
```

ðŸŽ¬ Join two data frames together using the combination of the treatment
and the antibody and calculate what % cells remained in each
sample after gating (*i.e.*, the % live cells):

```{r}
clean_trans_live_n <- clean_trans_live_n |> 
  left_join(clean_trans_n, by = c("antibody", "treatment")) |> 
  mutate(perc_live = round(n_live/n * 100, 1) )
```

```{r}
#| echo: false
knitr::kable(clean_trans_live_n)
```

ðŸŽ¬ Plot the forward scatter and side scatter for each sample with the
gate and the number of cells in each sample.

```{r}
ggplot(clean_trans, aes(x = FS_Lin, y = SS_Lin)) +
  geom_hex(bins = 128) +
  scale_fill_viridis_c() +
  geom_polygon(data = box, aes(x = x, y = y), 
               fill = NA, 
               color = "red",
               linewidth = 1) +
  geom_text(data = clean_trans_live_n, 
            aes(label = paste0(perc_live, "%")), 
            x = 45000, 
            y = 1000,
            colour = "red",
            size = 5) +
  facet_grid(antibody ~ treatment) +
  theme_bw() 



```

::: callout-tip
## Reporting

This figure allows you to report what data you included in your analysis
of the `TNFa_APC_Lin` and `E_coli_FITC_Lin` signals.
:::



ðŸŽ¬ Plot the `TNFa_APC_Lin` and `E_coli_FITC_Lin` signals for these live
cells:

```{r}
ggplot(clean_trans_live, aes(x = E_coli_FITC_Lin, 
                                  y = TNFa_APC_Lin)) +
  geom_hex(bins = 128) +
  scale_fill_viridis_c() +
  facet_grid(antibody ~ treatment) +
  theme_bw()
```

## Quality control 3: Gating to determine a 'real' signal?

From now on we are working with the cleaned and gated data, that is,
the dataframe `clean_trans_live`.

We have two signals:

-   `TNFa_APC_Lin` is should indicate the amount of TNF-Î± protein in the
    cell
-   `E_coli_FITC_Lin` should indicate the amount of *E. coli* in the
    cell

### The `TNFa_APC_Lin` signal

When the antibody is ISOTYPE there is no TNF-Î± so that level of
`TNFa_APC_Lin` signal means no TNF-Î±. In other words, that signal is a 
control. If you look at the top row of scatter plots, that level seems 
to be about 3.8. In the rest of the data set we should assume any signal
below 3.8 means there is no TNF-Î±.

We can use this to set a threshold for the `TNFa_APC_Lin` signal and
then label cells as either positive or negative for TNF-Î±.

ðŸŽ¬ Define the threshold for the `TNFa_APC_Lin` signal:

```{r}
apc_cut <- 3.8
```

ðŸŽ¬ Plot the `TNFa_APC_Lin` and `E_coli_FITC_Lin` with the threshold for
the `TNFa_APC_Lin` signal:

```{r}
ggplot(clean_trans_live, aes(x = E_coli_FITC_Lin, 
                                  y = TNFa_APC_Lin)) +
  geom_hex(bins = 128) +
  geom_hline(yintercept = apc_cut, 
             color = "red") +
  scale_fill_viridis_c() +
  facet_grid(antibody ~ treatment) +
  theme_bw()
```

That looks about right. We will use this value to label each cell as 
positive or negative for TNF-Î±.

ðŸŽ¬ Add a label, `tnfa`, to the data to indicate if the cell is positive
or negative for TNF-Î±

```{r}
clean_trans_live <- clean_trans_live |> 
  mutate(tnfa = case_when(TNFa_APC_Lin < apc_cut ~ "TNF-Î± -'ve",
                               TNFa_APC_Lin >= apc_cut ~ "TNF-Î± +'ve"))
```

ðŸŽ¬ Summarise each group by finding the number of TNF-Î± positive cells
and the mean `TNFa_APC_Lin` signal for each group:

```{r}
clean_trans_live_tfna_pos <- clean_trans_live |> 
  filter(tnfa == "TNF-Î± +'ve") |>
  group_by(antibody, treatment) |>
  summarise(n_pos_tnfa = n(),
            mean_apc = mean(TNFa_APC_Lin))
```

```{r}
#| echo: false
knitr::kable(clean_trans_live_tfna_pos)
```

ðŸŽ¬ We can add the number of live cells to this data frame and
calculate the percentage of live cells that are positive for
TNF-Î±:

```{r}
clean_trans_live_tfna_pos <- 
  clean_trans_live_tfna_pos |> 
  left_join(clean_trans_live_n, by = c("antibody", "treatment")) |> 
  mutate(perc_pos_tnfa = round(n_pos_tnfa/n_live * 100, 1) )
```


ðŸŽ¬ And put the columns into a more logical order:

```{r}
clean_trans_live_tfna_pos <- 
  clean_trans_live_tfna_pos |> 
  select(antibody,
         treatment,
         n,
         n_live,
         perc_live,
         mean_apc,
         n_pos_tnfa,
         perc_pos_tnfa)

```



```{r}
#| echo: false
knitr::kable(clean_trans_live_tfna_pos)
```

The fact that a very low percentage of cells are positive for TNF-Î± in
the isotype control is a good indication that cut off we used for the
`TNFa_APC_Lin` signal is a good one.

TNFAPC-MEDIA combination tells us how much TNF-Î± we expect in 
unstimulated cells.

This is important data for your write up and will be your contribution 
to the calss data set so you should write it to file.

ðŸŽ¬ Write `clean_trans_live_tfna_pos` to file:

```{r}
write_csv(clean_trans_live_tfna_pos, 
          "data-processed/clean_trans_live_tfna_pos.csv")
```

### The `E_coli_FITC_Lin` signal

We can apply the same logic to the `E_coli_FITC_Lin` signal.

For either antibody, when the treatment is media or LPS there is no
*E.coli* so that level of `E_coli_FITC_Lin` signal means zero. That 
level looks to be about 2.

ðŸŽ¬ Define the threshold for the `E_coli_FITC_Lin` signal:

```{r}
fitc_cut <- 2
```

ðŸŽ¬ Add a label, `fitc`, to the data to indicate if the cell is positive
or negative for FITC

```{r}
clean_trans_live <- clean_trans_live |> 
  mutate(fitc = case_when(E_coli_FITC_Lin < fitc_cut ~ "FITC -'ve",
                               E_coli_FITC_Lin >= fitc_cut ~ "FITC +'ve"))
```

::: callout-tip
## Tip for your own summaries

You may wish to repeat the process we used for the `TNFa_APC_Lin`
signal to find out the number of cells that are FITC positive in each 
sample and the mean intensity of the FITC signal in the FITC positive 
cells.
:::

### A figure to report baselines for both signals

ðŸŽ¬ Plot the TNFa_APC_Lin and E_coli_FITC_Lin with the thresholds for
both signals:

```{r}
ggplot(clean_trans_live, aes(x = E_coli_FITC_Lin, 
                                  y = TNFa_APC_Lin)) +
  geom_hex(bins = 128) +
  geom_hline(yintercept = apc_cut, 
             color = "red") +
    geom_vline(xintercept = fitc_cut, 
             color = "red") +
  scale_fill_viridis_c() +
  facet_grid(antibody ~ treatment) +
  theme_bw()
```



## Save the data

We have filtered out the dead cells and debris and added labels to the
data to indicate if the cells are positive or negative for TNF-Î± and *E.
coli*. It would be a good idea to save this data to file so we can start
from importing this data in the future rather than having to repeat all
of the steps we have done so far.

ðŸŽ¬ Save the data:

```{r}
write_csv(clean_trans_live, "data-processed/live_labelled.csv")

```

## Look after future you!

ðŸŽ¬ This workshop is a template for the analysis of your own data. Go
back through your script and check you understand what you have done.
Can you identify where you will need to edit? Do you understand what you
will need to assess in order to make edits needed for your own data?
Examine all of the objects in the environment to check you understand
what they are and why they are there.

You're finished!

# ðŸ¥³ Well Done! ðŸŽ‰

# Independent study following the workshop

[Consolidate](study_after_workshop.qmd)

# The Code file

This contains all the code needed in the workshop even where it is not
visible on the webpage.

The `workshop.qmd` file is the file I use to compile the practical. Qmd
stands for Quarto markdown. It allows code and ordinary text to be
interweaved to produce well-formatted reports including webpages. [View
the Qmd in
Browser](https://github.com/3mmaRand/R4BABS/blob/main/r4babs4/week-2/workshop.qmd).
Coding and thinking answers are marked with `#---CODING ANSWER---` and
`#---THINKING ANSWER---`

Pages made with R [@R-core], Quarto [@allaire2022], `knitr` 
[@knitr1; @knitr2; @knitr3],
`kableExtra` [@kableExtra]

# References
