---
title: "Workshop"
subtitle: "Data Analysis 4: Immunobiology - Your data presentation"
toc: true
toc-location: right
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| include: false
library(tidyverse)
library(kableExtra)
```

In this workshop 

# Exercises

## Add your data

Enter these in the [BIO00066I Biomedical Sciences class data](https://docs.google.com/spreadsheets/d/104EXdgsiIq-FuRF9Ly9zewEVdpkVWbyOwxSAmiqJepg/edit#gid=0)

The columns you must add are:
-   `apc_mfi`:	Mean fluorescence intensity of the logicle transformed TNFa_APC_Lin in the TNF-α positive cells

-   `perc_tfna_pos`:	% non debris cells that are TNF-α positive cells   

The other columns are calculations you make along the way and may help you get 
to the `apc_mfi` and `perc_tfna_pos` values. The column names are the same as 
those used in the [Data Analysis 2: Immunobiology - Sample data analysis](../week-2/workshop.html) workshop

## Organising your work

script for import and cleaning and writing clean data to file because time consuming

script for importing cleaned data gating live cells, summarising and plotting



## Importing from google sheets

The [BIO00066I Biomedical Sciences class data](https://docs.google.com/spreadsheets/d/104EXdgsiIq-FuRF9Ly9zewEVdpkVWbyOwxSAmiqJepg/edit#gid=0) are in a google sheet.
You can download the file as an excel file for .csv.
file but it is also possible to import the data directly from the google sheet into R.
Use whatever you prefer.
An advantage of using the google sheet is you won't have to remember to download the data when someone updates it.

You can use the `googlesheets4` package [@googlesheets4] to do this.

```{r}
library(googlesheets4)
```

## Import cleaned data


gating live cells


# Independent study following the workshop

[Consolidate](study_after_workshop.qmd)

# The Code file

This contains all the code needed in the workshop even where it is not visible on the webpage.

The `workshop.qmd` file is the file I use to compile the practical.
Qmd stands for Quarto markdown.
It allows code and ordinary text to be interweaved to produce well-formatted reports including webpages.
[View the Qmd in Browser](https://github.com/3mmaRand/R4BABS/blob/main/r4babs4/week-6/workshop.qmd).
Coding and thinking answers are marked with `#---CODING ANSWER---` and `#---THINKING ANSWER---`

Pages made with R [@R-core], Quarto [@allaire2022], `knitr` [@knitr], `kableExtra` [@kableExtra]

# References
