---
title: "Week 3: Two-sample tests"
subtitle: "Independent Study to consolidate this week"
format:
  html:
    code-fold: true
    code-summary: "Answer - don't look until you have tried!"
---

# Set up

If you have just opened RStudio you will want to load the **`tidyverse`** package

```{r}
#| code-fold: false
library(tidyverse)
```

# Exercises

1.  💻 Plant Biotech. Some plant biotechnologists are trying to increase the quantity of omega 3 fatty acids in Cannabis sativa. They have developed a genetically modified line using genes from *Linum usitatissimum* (linseed). They grow 50 wild type and fifty modified plants to maturity, collect the seeds and determine the amount of omega 3 fatty acids. The data are in [csativa.txt](data-raw/). Do you think their modification has been successful?


```{r}
#| output: false

# Unedited c&P

#---THINKING AND CODING ANSWER---
csativa  <-  read_table("../data/csativa.txt")
str(csativa)

# First realise that this is a two sample test. You have two independent samples
#  - there are a total of 100 different plants and the values in one 
#  group have no relationship to the values in the other.

# create a rough plot of the data  
ggplot(data = csativa, aes(x = plant, y = omega)) +
  geom_violin()
# note the modified plants seem to have lowere omega!

# create a summary of the data
csativasum <- csativa %>%
  group_by(plant) %>%
  summarise(mean = mean(omega),
            std = sd(omega),
            n = length(omega),
            se = std/sqrt(n))

# The data seem to be continuous so it is likely that a t-test will be fine
t.test(data = csativa, omega ~ plant, var.equal = TRUE)
# So there is a significant difference but you need to make sure you know the direction!

# let's check the assumptions
# add the group means to the data
csativa <- merge(csativa, csativasum[,1:2], by = "plant")
# add the residuals
csativa <- csativa %>%
  mutate(residual = omega - mean)

# plot
ggplot(data = csativa,
       aes(x = mean, y = residual)) +
  geom_point()
# normality test
shapiro.test(csativa$residual)

# One could argue that the variance is greater in one group. In which case you could run a Welch's t-test
t.test(data = csativa, omega ~ plant)

# You would write it up like this:
# Wild plants have a significantly higher omega 3 content (($\bar{x} \pm s.e.$ =  56.41 $\pm$ 1.11) than modified plants (49.46 $\pm$ 0.82)(Welch's t-test: t = 5.03; _d.f._ = 90.36; _n1_ =  50, _n2_ = 50; _p_ < 0.001).

# A figure 
ggplot() +
  geom_point(data = csativa, aes(x = plant, y = omega),
             position = position_jitter(width = 0.1, height = 0),
             colour = "gray50") +
  geom_errorbar(data = csativasum, 
                aes(x = plant, ymin = mean - se, ymax = mean + se),
                width = 0.3) +
  geom_errorbar(data = csativasum, 
                aes(x = plant, ymin = mean, ymax = mean),
                width = 0.2) +
  scale_x_discrete(name = "Plant type", labels = c("GMO", "WT")) +
  scale_y_continuous(name = "Amount of Omega 3 (units)",
                     expand = c(0, 0),
                     limits = c(0, 80)) +
  theme_classic()

```

1.  💻 another example