---
title: "Independent Study - after the workshop"
format:
  html:
    code-fold: true
    code-summary: "Answer - don't look until you have tried!"
---

# Set up

If you have just opened RStudio you will want to load the packages and import the data.

```{r}
#| code-fold: false
library(tidyverse)
library(readxl)
```

```{r}
#| code-fold: false
fly_bristles_means <- read_excel("data-raw/bristles-mean.xlsx")
cats <- read_csv("data-raw/cat-coats.csv")
```

# Exercises

1.  💻 Summarise the `fly_bristles_means` dataframe by calculating the mean, median, sample size, standard deviation and standard error of the `mean_count` variable.

```{r}
#| code-fold: true
fly_bristles_means |> 
  summarise(mean = mean(mean_count),
            median = median(mean_count),
            n = length(mean_count),
            standard_dev = sd(mean_count),
            standard_error = standard_dev / sqrt(n))
```

2.  💻 Create an appropriate plot to show the distribution of `mean_count` in `fly_bristles_means`

```{r}
#| code-fold: true

ggplot(fly_bristles_means, aes(x = mean_count)) +
  geom_histogram(bins = 10)
```

3.  💻 Can you format the plot 2. by removing the grey background, giving the bars a black outline and the fill colour of your choice and improving the axis format and labelling? You may want to refer to [last week's workshop](../week-7/workshop.html).


```{r}
#| code-fold: true

ggplot(fly_bristles_means, aes(x = mean_count)) +
  geom_histogram(bins = 10, 
                 colour = "black",
                 fill = "skyblue") +
  scale_x_continuous(name = "Number of bristles",
                     expand = c(0, 0)) +
  scale_y_continuous(name = "Frequency",
                     expand = c(0, 0),
                     limits = c(0, 35)) +
  theme_classic()
```

4.  💻 Amend this code to change the order of the bars by the average mass of each coat colour? [Changing the order of bars](../week-7/workshop.html#changing-the-order-of-bars) was covered last week. You may also want to practice formatting the graph nicely.

```{r}
#| echo: true
#| include: true

ggplot(cats, aes(x = coat, y = mass)) +
  geom_boxplot()
```

```{r}
#| code-fold: true
ggplot(cats, 
       aes(x = reorder(coat, mass), y = mass)) +
  geom_boxplot(fill = "darkcyan") +
  scale_x_discrete(name = "Coat colour") +
  scale_y_continuous(name = "Mass (kg)", 
                     expand = c(0, 0),
                     limits = c(0, 8)) +
  theme_classic()
```

4.  📖 Read [Understanding the pipe `|>`](https://3mmarand.github.io/comp4biosci/workflow_rstudio.html)
