---
title: "Independent Study to consolidate this week"
subtitle: "Association: Correlation and Contingency"
toc: true
toc-location: right
format:
  html:
    code-fold: true
    code-summary: "Answer - don't look until you have tried!"
---

# Set up

If you have just opened RStudio you will want to load the **`tidyverse`** package

```{r}
#| code-fold: false
library(tidyverse)
```

# Exercises

1.  💻 The slug *Arion ater* has three major colour forms, black, chocolate brown and red. Sampling in population X revealed 27 black, 17 brown and 9 red individuals, whereas in population Y the corresponding numbers were 39, 10 and 21. Create an appropriate data structure and test whether the proportion of black, brown and red slugs differs between the two populations.

```{r}
#| output: false

#  names for the rows and columns
vars <- list(pop = c("x","y"), 
             colour = c("black", "brown", "red"))

# matrix of the data with named columns and rows
slugs <- matrix(c(27, 39, 17, 10, 9, 21), 
                nrow = 2, dimnames = vars)
slugs
# gives me
#    colour
# pop black brown red
#   x    27    17   9
#   y    39    10  21

# you may need to try a couple of times to get the numbers in 
# the right places

```

```{r}
#| output: false

chisq.test(slugs)
# X-squared = 6.5726, df = 2, p-value = 0.03739

# p < 0.05 so we reject the null hypothesis i.e., the proportions of the colour 
# forms are significantly different in the two populations 
# (mostly as a result of differences in the brown and red classes - 
# look at the differences between observed and expected values for 
# the three colour forms in the table above).
chisq.test(slugs)$expected
#       colour
# pop    black    brown      red
#   x 28.43902 11.63415 12.92683
#   y 37.56098 15.36585 17.07317

```


2. 💻  The raw, untabulated data are in [slugs.txt](data-raw/slugs.txt). Perform the test on these data.

```{r}
#| output: false

# import the data
slugs <- read_table("data-raw/slugs.txt")

# put it into a table
slugtab <- table(slugs$colour, slugs$pop)

# carry out the test
chisq.test(slugtab)
```


